Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2019-02-18 Manuscript/Text/manuscript.tex
Most reviewer revisions

Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how low diversity diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.

Viral infection was a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteSpread}. This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family Dicistroviridae \cite{fam}. IAPV infection causes shivering wings, decreased locomotion, muscle spasms, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}. IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.

@@

Following up on these findings, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on vigor (tolerance) \cite{resTol1}. Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}.

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms. Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}. Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}. Studies of virus-bee systems have revealed some of the antiviral defense pathways known in model organisms are conserved and also related to bee antiviral immune responses \cite{McMenamin}.

To our knowledge, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}. In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). For our diet factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.). Rockrose pollen is generally considered less nutritious than chestnut pollen because it contains smaller amounts of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}. We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

Because RNA-seq data can be noisy and subject to high levels of inter-experiment variation, we further sought to validate our transcriptomic data via comparison to a previous RNA-seq study on honey bee responses to viral infection. To do this, we compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used honey bees from 15 different colonies, i.e. from 15 different, naturally-mated queens. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. To achieve this comparison, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

As shown in Figure 1, mortality rates of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02). Overall, we interpreted these findings to mean that high-quality chestnut pollen could partially ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).

@@

We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). In the diet factor, more DEGs were upregulated in the more-nutritious chestnut group (\textit{n} = 1,033) than in the less-nutritious rockrose group (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18).

GO analysis of the chestnut-upregulated DEGs revealed the following over-represented biological functions: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1). GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online \cite{bigpint}. To understand how to interpret the visualizations used in this section, please read (https://lrutter.github.io/bigPint/articles/plotIntro.html).

We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of \textit{all} genes in the data, and each parallel coordinate line represents one DEG. In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed. We expect DEGs to show more variability between treatments than between replicates. This means the parallel coordinate lines should be flat between replicates but crossed between treatments. However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot. Therefore, we graphed clustered subsets of the DEGs (based on hierarchical clustering).

The 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3). In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large (or even consistent) difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

@@

We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

@@

Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens \cite{AdamTothReview}. In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the prominent virus of concern, IAPV. In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet). The majority of these genes were upregulated in response to high quality diet, and these genes were over-represented for functions such as nutrient signaling metabolism (insulin resistance), immune response (Notch signaling and JaK-STAT pathways), and carbon metabolism (Supplementary table 2, Additional file 1). These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into innate immune responses.

@@

While some insect systems have shown relatively low transcriptional responses to dicistrovirus infection \cite{fewDEGs, fewDEGs2}, previous work on honey bees has revealed many hundreds of DEGs \cite{galbraith}. Discrepancies between datasets may be due to noise and complexity of the honey bee microbiome. The transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 differentially expressed transcripts, some with known immune functions such as a gene with similarity to MD-2 lipid recognition protein and argonaute-2, a protein that plays a central role in RNA silencing (Table \ref{tbl:virusGenes}). We also found genes related to transcriptional regulation and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14).

@@

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion (Figure 5B). Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

>>>>

Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how low diversity diets (i.e. monofloral diets) affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}. Indeed, differing qualities of monofloral diets have been shown to affect nurse bee physiology and tolerance to parasites \cite{DiPasquale}.

Viral infection was considered a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteProblem, miteProblem2, miteProblem3}. This mite feeds on honey bee hemolymph and/or fat body tissue \cite{hemolymph, fatbody}, and is believed to decrease lipid and glycogen reserves and reduce protein synthesis in bees \cite{miteNutrition}. Additionally, it transmits multiple viruses and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family \textit{Dicistroviridae} \cite{fam}. IAPV infection causes shivering wings, decreased locomotion, muscle spasms, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}. IAPV has demonstrated higher infectious capacity \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.

@@

Following up on these findings, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function that reduces the pathogen load of the host (resistance, \cite{Medzhitov}) or if it is due to indirect effects of good nutrition on the ability of the host to reduce pathogen impacts without affecting pathogen load (tolerance, \cite{Medzhitov}). Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection, which can help us generate candidate gene lists to further investigate the relative roles of tolerance and resistance. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}.

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith, Brutscher, Chen, Ryabov, Doublet} in honey bees, but there have been several such studies in model organisms. Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2, Chen}. Previous transcriptomic studies have identified genes serving links between metabolism and antiviral defense in honey bees \cite{metabolismVirus, metabolismVirus1}; see \cite{metabolismVirusReview} for an overview. Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}. Studies of virus-bee systems have revealed some of the antiviral defense pathways known in model organisms are conserved and also related to bee antiviral immune responses \cite{McMenamin}.

To our knowledge, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}. In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). For our diet factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.). Rockrose pollen is generally considered less nutritious than chestnut pollen because it contains smaller amounts of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}. For specific quantitative differences between these two pollen groups, please see \cite{DiPasquale}. We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

Because RNA-seq data can be noisy and subject to high levels of inter-experiment variation, we further sought to validate our transcriptomic data via comparison to a previous RNA-seq study on honey bee responses to viral infection. To do this, we compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used honey bees from 15 different colonies, i.e. from 15 different, naturally-mated queens. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions. To achieve this comparison, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches (including scatterplot matrices, parallel coordinate plots, and litre plots from the bigPint software package) can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

As shown in Figure 1, mortality rates of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Bees fed rockrose pollen had significantly elevated mortality with virus infection compared to non-inoculated controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to non-inoculated controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to non-inoculated controls (Benjamini Hochberg, p = 6.29e-02). While many of the non-inoculated treatment groups showed some RT-qPCR amplification (non-inoculated average Ct=33.92; inoculated average Ct=24.9), and thus have virus titers calculable on a standard curve, these Ct levels are similar to those deemed uninfected in previous studies \cite{galbraith}. Overall, we interpreted these findings to mean that high-quality chestnut pollen could partially reduce high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).

@@

We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). There were only four genes that were DEGs in both our diet main effect and our virus main effect (GB48747, GB47214, GB42908, and GB42507). In the diet factor, more DEGs were upregulated in the more-nutritious chestnut group (\textit{n} = 1,033) than in the less-nutritious rockrose group (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18). For our DEG analysis, we used R software version 3.3.3 \cite{R}.

We performed GO analysis to statistically assign our DEGs to predefined bins based on their functional characteristics, allowing us to better understand the biological processes of our DEGs. GO analysis of the chestnut-upregulated DEGs revealed the following over-represented biological functions: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1). GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three individual bees per treatment group as a single pooled sample, while our dataset contained 16 individual bees per treatment group in 8 RNA-seq samples. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and non-inoculated honey bees better than our dataset (Additional file 2). Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online \cite{bigpint}. For more information about the visualizations used here, please refer to  (https://lindsayrutter.github.io/bigPint/articles/plotIntro.html).

We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of \textit{all} genes in the data (all 15,314 genes in our count table), and each parallel coordinate line represents one DEG. In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed. We expect DEGs to show more variability between treatments than between replicates. This means the parallel coordinate lines should be flat between replicates but crossed between treatments. However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot. Therefore, we graphed clustered subsets of the DEGs (based on hierarchical clustering).

The 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments. The few inconsistent replicates we observed (such as V.1 from Cluster 1 and V.2 from Cluster 4) were small enough that consistent differences between the treatment groups remained apparent across the samples (Figure 3). In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large (or even consistent) difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

@@

We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). Specifically, we arbitrarily subsetted the samples so each one was represented once in each of these four files (i.e. Additional File 11 shows samples 1-3; Additional File 12 shows samples 4-6; Additional File 13 shows samples 7-9; and Additional File 14 shows samples 10-12). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

@@

Challenges to honey bee health are a growing concern, in particular the combined, interactive effects of nutritional stress and pathogens \cite{AdamTothReview}. In this study, we used RNA-sequencing to probe mechanisms underlying honey bee responses to two effects, diet quality and infection with the prominent virus of concern, IAPV. In general, we found a major nutritional transcriptomic response, with nearly 2,000 transcripts changing in response to diet quality (rockrose/poor diet versus chestnut/good diet). The majority of these genes were upregulated in response to high quality diet, and these genes were over-represented for functions such as nutrient signaling metabolism (insulin resistance), immune response (Notch signaling and JaK-STAT pathways), and carbon metabolism (Supplementary table 2, Additional file 1). These data suggest high quality nutrition may allow bees to alter their metabolism, favoring investment of energy into immune responses.

@@

While some insect systems have shown relatively low transcriptional responses to dicistrovirus infection \cite{fewDEGs, fewDEGs2}, previous work on honey bees has revealed many hundreds of DEGs \cite{galbraith}. Discrepancies between datasets may be due to noise and complexity of the honey bee microbiome. The transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 differentially expressed transcripts, some with known immune functions such as a gene with similarity to MD-2 lipid recognition protein and argonaute-2, a protein that plays a central role in RNA silencing (Table \ref{tbl:virusGenes}). We also found genes related to transcriptional regulation and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14) and baseline viral titers observed in our control bees (Figure 2).

@@

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion (Figure 5B). Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}. Thus, our study uses transcriptome data to generate lists of candidate genes that can be the focus of future investigations to better experimentally test putative roles of tolerance and resistance genes in this system.

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (naturally-mated colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on worker honey bees that were fed sugar and artificial pollen diets, whereas our study focused on worker honey bees that were fed bee-collected monofloral diets. Also, Galbraith’s bees were sampled at 24 hours while ours were sampled at 36 hours. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and observations to determine behaviorally symptomatic bees whereas we used whole bodies and categorized only into inoculated vs. non-inoculated groups. There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. The different factors between these two studies may be critical because particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}.
